Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-25 manuscript.rmd
1. edited symbols in SI to match with the manuscript

The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018]. The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO2/H2O gas analyzer (LI7500, LiCOR. Lincoln, NE). Hi-frequency (10Hz) measurements were acquired by a datalogger (CR1000, Campbell Scientific) and stored on a local PC. Data were processed with a custom program using a standard routine described in [@Detto.2010]. QA/QC criteria for removing erroneous values are listed in table 1 (see below). GPP was derived from daytime values of NEE by adding the corresponding mean daily ecosystem respiration obtained as the intercept of the light response curve [@Lasslop.2010]. The light curve was fitted on a 15-days moving window using a rectangular hyperbolic function (runs with friction velocity less than 0.4 m s^-1^ were excluded). In order to compute daily time integrated budgets, gaps were filled using Artificial Neural Network [@Papale.2003] with hydro-meteorological inputs as predictors (soil moisture, solar radiation, temperature, VPD and air pressure, all measured on the same tower). To train the network, the dataset was randomly divided in a training set (70%), a validation set (15%) and a test set (15%). A two-layer feed-forward network with 10 sigmoid hidden neurons and linear output neurons, was trained using the Levenberg-Marquardt algorithm until the mean square error (MSE) of the validation set stop improving [@Hagan.1994]. Performance, in term of MSE, was evaluated using the test set at the end of the training. This procedure was repeated 100 times to produce 100 estimates of GPP. Training multiple times generates different results due to different initial conditions and random sampling of the three sets. Ensemble was obtained as weighted average from the 100 ANN predictions using the MSE of the test set as weights according to:

@@

We also leveraged exisiting GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot. The sampling periods were 6, 5, 10 and 8 days long, respectively. The 2016 dry season was associated with the 2015–2016 El Niño (see @Kupers:2019 for further details). We converted these GWC measurements to VWC assuming a bulk density of 0.8 g cm^-3^. $\theta_{0.15 m}$, $\theta_{0.4 m}$ and $\theta_{1 m}$ refer to the plot-wide VWC averages by depth over each sampling period. Middle date of a sampling period is used for comparison with model simulations.

@@

Brett to edit.
\
Methods for Stem Maximum hydraulic conductivity and vulnerability to cavitation closely followed [Christman.2012] and [Wolfe.2016]. Values for some of the species reported here were previously reported by @Dickman.2019 and @Wu.2020. The entire dataset and R script for the analyses are archived in the NGEE-Tropics archive (Wolfe 2020, To be done).

\   Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX). The branch segments were enclosed in opaque plastic bags, and brought to the laboratory for measurements. Stem segments used for hydraulic conductivity (Kh) measurements were located at least 1 m from the cut end of the branch to minimize the number of open vessels that extended into the segment. Maximum vessel length was measured on a subset of samples from each species. Among species, maximum vessel length ranged XXX and length of stem between the cut that was made in the air to collect the branch and the section of stem used for Kh measurement was XXX (Table SXX).

\   Three leaves that were distal to the Kh segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration. Just prior to measurement of Kh, leaf water potential of the three bagged leaves was measured with a pressure chamber. Since transpiration was stopped by bagging the leaves, leaf water potential was assumed to equal stem water potential (REFERENCE).

\   For each Kh segment, three measurements of stem water potential were made on stem sections that were located adjacent to the Kh segment. Stem sections that were approximately 40 mm in length were placed in nickel-plated aluminum chambers that were connected to Peltier-type psychrometers (Merrill Instruments, Logan, UT) and placed in a water bath at 30°C for 3 h to equilibrate the humidity within the cambers. Then water potential was measured using an automated multi-channel micro-voltmeter (CR7, Campbell Scientific, Logan, UT) with a 10 s thermocouple cooling current. For samples with water potential < -4 MPa, a 45 s cooling current was used to improve accuracy. Each psychrometer probe was calibrated against salt solutions beforehand and the plateau of the psychromatic response curve was used to assess water potential.

\   The Kh segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)). Then the ends were shaved clean with a fresh razor. Segments were 11–34 cm in length (mean ± SD = 22 ± 4 cm) and 0.2–1.6 cm in diameter (0.7 ± 0.2 cm). The segments were placed in a hydraulic conductivity apparatus similar to that described by Sperry et al. (1998) except that flow rate was measured with graduated pipettes instead of a balance. The perfusion solution was 10 mM KCl filtered to 0.2 m. For each segment, we measured the flow rate at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure. Following Torres-Ruiz et al. (2012), Kh was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads. Stem area specific hydraulic conductivity was (Ks) was calculated by dividing Kh by the cross-sectional area of stem. To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.

\   To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and Ks. We used the water potential measured by the psychrometers because they are a more direct measurement. It has been noted that bagged-leaf water potential can deviate from stem water potential, especially when branch segments are relatively dry (REFERENCE). For some species, we noted this pattern while for others it was not apparent (Fig SXX). For each species, Ks was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al. (2016)). The intercept of the curve was interpreted as maximum Ks and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum Ks were interpreted as P12, P50, P80, and P88, respectively.

@@

Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs. \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~. We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj. $R$^2^ = `r disc.r2`; $p$ `r disc.p`, $n$ = `r disc.n`). We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj. $R$^2^ = `r lma.r2`; $p$ `r lma.p`, $n$ = `r lma.n`). We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.

>>>>

The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018]. The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO~2~/H~2~O gas analyzer (LI7500, LiCOR. Lincoln, NE). Hi-frequency (10Hz) measurements were acquired by a datalogger (CR1000, Campbell Scientific) and stored on a local PC. Data were processed with a custom program using a standard routine described in [@Detto.2010]. QA/QC criteria for removing erroneous values are listed in table 1 (see below). GPP was derived from daytime values of NEE by adding the corresponding mean daily ecosystem respiration obtained as the intercept of the light response curve [@Lasslop.2010]. The light curve was fitted on a 15-days moving window using a rectangular hyperbolic function (runs with friction velocity less than 0.4 m s^-1^ were excluded). In order to compute daily time integrated budgets, gaps were filled using Artificial Neural Network [@Papale.2003] with hydro-meteorological inputs as predictors (soil moisture, solar radiation, temperature, VPD and air pressure, all measured on the same tower). To train the network, the dataset was randomly divided in a training set (70%), a validation set (15%) and a test set (15%). A two-layer feed-forward network with 10 sigmoid hidden neurons and linear output neurons, was trained using the Levenberg-Marquardt algorithm until the mean square error (MSE) of the validation set stop improving [@Hagan.1994]. Performance, in term of MSE, was evaluated using the test set at the end of the training. This procedure was repeated 100 times to produce 100 estimates of GPP. Training multiple times generates different results due to different initial conditions and random sampling of the three sets. Ensemble was obtained as weighted average from the 100 ANN predictions using the MSE of the test set as weights according to:

@@

We also leveraged exisiting GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot. The sampling periods were 6, 5, 10 and 8 days long, respectively. The 2016 dry season was associated with the 2015–2016 El Niño (see @Kupers:2019 for further details). We converted these GWC measurements to VWC assuming a bulk density of 0.8 g cm^-3^. $\theta_{\textrm{0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to the plot-wide VWC averages by depth over each sampling period. Middle date of a sampling period is used for comparison with model simulations.

@@

Brett to edit.
\
Methods for Stem Maximum hydraulic conductivity and vulnerability to cavitation closely followed [Christman.2012] and [Wolfe.2016]. Values for some of the species reported here were previously reported by @Dickman.2019 and @Wu.2020. The entire dataset and *R* script for the analyses are archived in the NGEE-Tropics archive (Wolfe 2020, To be done).

\   Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX). The branch segments were enclosed in opaque plastic bags, and brought to the laboratory for measurements. Stem segments used for hydraulic conductivity ($K_h$) measurements were located at least 1 m from the cut end of the branch to minimize the number of open vessels that extended into the segment. Maximum vessel length was measured on a subset of samples from each species. Among species, maximum vessel length ranged XXX and length of stem between the cut that was made in the air to collect the branch and the section of stem used for $K_h$ measurement was XXX (Table SXX).

\   Three leaves that were distal to the $K_h$ segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration. Just prior to measurement of $K_h$, leaf water potential of the three bagged leaves was measured with a pressure chamber. Since transpiration was stopped by bagging the leaves, leaf water potential was assumed to equal stem water potential (REFERENCE).

\   For each $K_h$ segment, three measurements of stem water potential were made on stem sections that were located adjacent to the $K_h$ segment. Stem sections that were approximately 40 mm in length were placed in nickel-plated aluminum chambers that were connected to Peltier-type psychrometers (Merrill Instruments, Logan, UT) and placed in a water bath at 30°C for 3 h to equilibrate the humidity within the cambers. Then water potential was measured using an automated multi-channel micro-voltmeter (CR7, Campbell Scientific, Logan, UT) with a 10 s thermocouple cooling current. For samples with water potential < -4 MPa, a 45 s cooling current was used to improve accuracy. Each psychrometer probe was calibrated against salt solutions beforehand and the plateau of the psychromatic response curve was used to assess water potential.

\   The $K_h$ segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)). Then the ends were shaved clean with a fresh razor. Segments were 11–34 cm in length (mean ± SD = 22 ± 4 cm) and 0.2–1.6 cm in diameter (0.7 ± 0.2 cm). The segments were placed in a hydraulic conductivity apparatus similar to that described by Sperry et al. (1998) except that flow rate was measured with graduated pipettes instead of a balance. The perfusion solution was 10 mM KCl filtered to 0.2 m. For each segment, we measured the flow rate at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure. Following Torres-Ruiz et al. (2012), $K_h$ was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads. Stem area specific hydraulic conductivity was ($K_stem$) was calculated by dividing $K_h$ by the cross-sectional area of stem. To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.

\   To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_stem$. We used the water potential measured by the psychrometers because they are a more direct measurement. It has been noted that bagged-leaf water potential can deviate from stem water potential, especially when branch segments are relatively dry (REFERENCE). For some species, we noted this pattern while for others it was not apparent (Fig SXX). For each species, $K_stem$ was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al. (2016)). The intercept of the curve was interpreted as maximum $K_stem$ and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum $K_stem$ were interpreted as $\\mathrm{\\Psi}$~12,stem~, $\\mathrm{\\Psi}$~50,stem~, $\\mathrm{\\Psi}$~80,stem~, and $\\mathrm{\\Psi}$~88,stem~, respectively.

@@

Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs. \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~. We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{\textrm{LMA}} = `r disc.cf[1]` + `r disc.cf[2]`\textrm{LMA}_{disc}$; Adj. $R$^2^ = `r disc.r2`; $p$ `r disc.p`, $n$ = `r disc.n`). We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{\textrm{LMA}} = `r lma.cf[1]` + `r lma.cf[2]`\textrm{LMA}_{lamina}$; Adj. $R$^2^ = `r lma.r2`; $p$ `r lma.p`, $n$ = `r lma.n`). We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.
